Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-24 paper.tex
section caterpillar convexity

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := [\hat T]$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$
		\WHILE {$\rank(\mrca_{\hat{T}}(C_i))>i$}
			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C_i))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

In this section we analyse shortest paths in $\rnni$.
Therefore, we will use the algorithms introduced in the previous section.
They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.

This algorithm gives an upper bound for the diameter of $\rnni$, which
\todo{Why is that?}
improves
the one given in \autocite{Gavryushkin2018-ol}.
\todo{When}
Later we will see that this bound is tight.
However, we don't know how good $\findpath$ approximates distances in $\rnni$. \todo{Can we say something about the approximation ratio?}

\todo{make sure trees are consistently denoted by upper case letters}
\begin{lemma}
$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$, where $\Delta(\rnni)$ denotes the diameter of the $\rnni$ graph, that is, if $d(T, R)$ is the graph distance in $\rnni$ then $\Delta(\rnni) = \max \{d(T, R) \mid T, R \in \rnni\}$.
\label{lemma:diameter_bound}
\end{lemma}

\begin{proof}
Since the diameter is bounded from above the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C$ induced by internal nodes of $R$ to be moved by $n - 1 - i$ $\rnni$ operations in $\hat{T}$.
Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths using an algorithm to compute shortest paths within the set of caterpillar trees and comparing them with shortest path in $\rnni$.
This algorithm will be explained in detail later in this section, after demonstrating why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.

We now compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move resolves it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
Therefore, a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
	\caption{Paths between caterpillar trees $T$ and $R$: black -- shortest path in classical $\nni$ (ignoring ranks) of length $5$; red -- shortest path in $\rnni$ of length $6$; blue -- path in $\rnni$ which is the natural extension of the shortest path in $\nni$ but has length $8$.}
	\label{fig:NNI_vs_RNNI}
\end{figure}

The above observation suggests that there is always a caterpillar path that is shortest path between two arbitrary caterpillar trees in $\rnni$.
Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that we by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.

@@

    For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
    We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:
    Without loss of generality we can assume that $T$ is the caterpillar tree $(\ldots ((1,2),3), \ldots, n)$ or $(\ldots ((1,2),3), \ldots, n+1)$ in the induction step.
    Let $R$ be a caterpillar tree with distance $\frac{(n-1)(n-2)}{2}$ to $T$.

    Induction basis: $n=3$\\
    There are three trees on $3$ taxa that have distance $1$ from another and are caterpillar trees.
    Hence it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For trees $T, R$ on $n$ taxa it is $d(T, R) = \frac{(n-1)(n-2)}{2}$.\\
    Let $T, R$ be trees on $n+1$ taxa with caterpillar distance $\frac{n(n-1)}{2}$.
    Therefore, each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
    From this we can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$. \todo{Do we need to explain this further (figure)?}
    Let us consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    We also know $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = n-1$.
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
\end{proof}

@@

\begin{proof}
    We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
    Since we use backwards induction, we need to consider the maximum distance between any two caterpillar trees as base case, which is $\frac{(n-1)(n-2)}{2}$ as we saw above.

@@

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
    Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
    We consider the shortest caterpillar path computed by $\csort$.
    Since $l < \frac{(n-1)(n-2)}{2}$, there is one loop $k$ when running $\csort$ where less than $n-2-k$ taxon pairs swap positions.
    We assume that this loop number $k$ is the first one in which this occurs.
    This situation is illustrated in the top of Figure~\ref{fig:cat_convex_seq_example}.
    Whenever the number of swaps of taxa or, in terms of sequences, neighboured elements in the sequence is maximum in a loop, one of the two first elements swaps position with $n-2-k$ subsequent elements of the sequence.
    Since we assume that the number of swaps in loop $k$ is not maximal, this does not happen.
    Let us assume that $a$ is the first element that is compared with its left neighbour in loop $k$ but does not swap with it.
    Then the left neighbour $b$ of $a$ in $T$ is below $a$ in $R$ as well.
    Therefore, the tree $T'$ which can be received from $T$ by exchanging $a$ and $b$ must have caterpillar distance $d_c(T',R) = d_c(T,R) + 1 = l + 1$ to $R$.
    For an illustration of the $k^{th}$ loop when performing $\csort$ between $T$ and $R$ or $T'$ and $R$, respectively, consider Figure~\ref{fig:cat_convex_seq_example}.

    \begin{figure}[H]
    	\centering
    	\includegraphics[width=\textwidth]{cat_convex_seq_example}
    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ as described in proof of Theorem~\ref{thm:caterpillar_convex}.
        The trees are represented as sequences that are being ordered by $\csort$.
        Right before loop number $k$, $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
    	\label{fig:cat_convex_seq_example}
    \end{figure}

>>>>

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := [\hat T]$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$ \label{alg:find_path:line:_cluster}
		\WHILE {$\rank(\mrca_{\hat{T}}(C_i))>i$}
			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C_i))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

In this section we analyse shortest paths in $\rnni$.
Therefore, we will use the algorithms introduced in the previous section.
They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
In Section~\ref{section:diameter} we will investigate diameter and radius of the $\rnni$ graph, before we discuss a conjecture, the so-called Split Theorem, which was raised in \autocite{Gavryushkin2018-ol}.
We will give a counterexample to disprove their version of the Split Theorem and claim an alternative version.

Before we get into the details of these geometric properties of $\rnni$, we need to prove some statements that will be used a few times within this geometry section.
At first we in Lemma~\ref{lemma:diameter_bound} an upper bound of the \emph{diameter} $\Delta(\rnni)$ of the $\rnni$ graph, which is defined as $\Delta(\RNNI) = \max \{d(T, R) \mid T, R \in \rnni\}$.

\begin{lemma}
$\Delta(\rnni) \leq \frac{(n-1)(n-2)}{2}$ for $n \geq 3$.
\label{lemma:diameter_bound}
\end{lemma}

\begin{proof}
Since the diameter is bounded from above by the maximum length of a path computed by $\findpath$, it is enough to find this maximum.
The worst-case execution of $\findpath$ would require each of the $n-2$ clusters $C_i$, considered in Line~\ref{alg:find_path:line:_cluster} of Algorithm~\ref{alg:find_path}, to be moved by $n-1-i$ $\rnni$ operations in $\hat{T}$.
Hence, the maximum length of such a path is bounded by $\sum\limits_{i = 1}^{n-2} i = \frac{(n-2)(n-1)}{2}$.
\end{proof}

@@

In this section we restrict our attention to the set of \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
This type of tree is of particular interest in $\rnni$ because shortest paths between such trees in $\rnni$ differ from those in classical $\nni$ space.
This in combination with the importance of caterpillar trees in the proof of $\np$-hardness of $\nni$ distances in \autocite{Dasgupta2000-xa} suggests that understanding the geometry of paths between caterpillar trees helps investigating the complexity of computing $\rnni$ distances.
In particular, we are going to prove that one can always find a shortest path between two caterpillar trees that contains caterpillar trees only, meaning that the set of caterpillar trees is convex in $\rnni$.
We discovered this by computing caterpillar paths with $\csort$ and comparing them with shortest path in $\rnni$, computed by the exact algorithm as explained in Section~\ref{section:algorithm}.
In the following we demonstrate why finding shortest paths between caterpillar trees builds the basis for understanding the difference between $\nni$ and $\rnni$.

Therefore, we compare shortest paths between caterpillar trees in $\nni$ with those in $\rnni$.
In $\nni$, building a cherry and moving it may give a path that is shorter than a path where the taxa move separately in as it is illustrated out in Figure~\ref{fig:NNI_vs_RNNI}.
There is only one move needed to build and resolve a cherry, respectively, while the number of moves for moving the cherry is the same as for moving a single taxon in $\nni$.
But in $\rnni$ space there are additional rank moves needed to change the rank of the internal node of the cherry before an $\nni$ move resolves it.
This is due to the fact that $\nni$ moves are only allowed on edges of length one in $\rnni$.
Hence a path where taxa move separately and no second cherry is built, is shortest in $\rnni$.

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
    \vspace{2pt}
    \todo{LC: Change caption}
	\caption{Paths between caterpillar trees $T$ and $R$: black -- shortest path in classical $\nni$ (ignoring ranks) of length $5$; red -- shortest path in $\rnni$ of length $6$; blue -- path in $\rnni$ which is the natural extension of the shortest path in $\nni$ but has length $8$.}
	\label{fig:NNI_vs_RNNI}
\end{figure}

The above observation suggests that there is always a caterpillar path that is a shortest path between any two caterpillar trees in $\rnni$.
Indeed, we are going to prove Theorem~\ref{thm:caterpillar_convex} within this section, which states that the set of caterpillar trees is convex in $\rnni$.
Note that this is not true for the $\nni$ space, which is clear from the example in Figure~\ref{fig:NNI_vs_RNNI}.

Lemma~\ref{lemma:caterpillar_dist=diameter} regards caterpillar trees with distance $\frac{(n-1)(n-2)}{2}$ and will be needed for proving the convexity of the set of caterpillar trees in $\rnni$.
Notice that by Lemma~\ref{lemma:diameter_bound} $\frac{(n-1)(n-2)}{2}$ is the upper bound for the diameter of $\rnni$.

@@

    For proving the other direction of the statement, let now $T$ and $R$ have caterpillar distance $d_c(T,R) =  \frac{(n-1)(n-2)}{2}$.
    We prove $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$ of $T$ and $R$:

    Induction basis: $n=3$\\
    There exist three trees on three taxa: $[1,2,3], [1,3,2]$ and $[2,3,1]$.
    \todo{LC: Do we use list representation here? If so, we need to mention it}
    They all have distance one from another and are caterpillar trees.
    So it is $d(T,R) = d_c(T,R) = 1$ for all pairs of trees $T,R$ on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For caterpillar trees $T, R$ on $n$ taxa with caterpillar distance $d_c(T,R) = \frac{(n-1)(n-2)}{2}$ it is $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Without loss of generality we can assume that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$.
    Let $R$ be a caterpillar tree with distance $\frac{n(n-1)}{2}$ to $T$.
    Then each of the $\frac{n(n-1)}{2}$ pairs of taxa that is considered when running $\csort$ to compute a path $p$ from $R$ to $T$ exchanges.
    In particular, taxon $n+1$ exchanges with a taxon in each of the $n+1$ loops in $\csort$.
    \todo{LC: Is this explanation sufficient?}
    We can conclude that taxon $n+1$, whose parent has highest rank in $T$, is part of the cherry of $R$.
    Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
    $\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same exchanges of taxa as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
    Hence, the caterpillar distance is $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
    So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.

    According to Lemma~\ref{lemma:distance_delete_taxon}, the distance between $T{\big|}_n$ and $R{\big|}_n$ is at least $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
    We also know $|\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = n-1$.
    Suppose, contrary to our claim, that there is a path from $T$ to $R$ of length less than $\frac{n(n-1)}{2}$.
    By the above, it follows that $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| < \frac{n(n-1)}{2} - (n-1) < \frac{(n-1)(n-2)}{2}$.
    Since this is a contradiction to $d(T{\big|}_n, R{\big|}_n) = \frac{(n-1)(n-2)}{2}$, it is proven that $d(T,R) = \frac{n(n-1)}{2} $.
\end{proof}

@@

\begin{proof}
    We prove this Theorem by backwards induction on the length $l$ of a shortest caterpillar path between two caterpillar trees $T$ and $R$.
    Since we use backwards induction, the maximum distance between any two caterpillar trees, which is $\frac{(n-1)(n-2)}{2}$ (Lemma~\ref{lemma:diameter_bound}), is the base case.

@@

    Induction step: $l+1 \to l$\\
    Induction hypothesis: For each pair of caterpillar trees with caterpillar distance $l+1$ there is a shortest path between these trees that only consists of caterpillar trees, which means that the distance between these trees is $l+1$.
    Let $T$ and $R$ be caterpillar trees with caterpillar distance $l$.
    We consider the shortest caterpillar path computed by $\csort$.
    Since $l < \frac{(n-1)(n-2)}{2}$, there must be one of the $n$ loops in $\csort$ where less than $n-2-k$ taxon pairs swap positions.
    Let us consider loop number $k$ where this happens for the first time.
    This situation is illustrated in the top of Figure~\ref{fig:cat_convex_seq_example}.
    Whenever the number of swaps of taxa or, in terms of lists, neighboured elements in the list is maximum in a loop, one of the first two elements swaps position with $n-2-k$ subsequent elements of the list.
    Since we assume that the number of swaps in loop $k$ is not maximal, there are less than $n-2-k$ such moves.
    Let us assume that $a$ is the first element that is compared with its left neighbour in loop $k$ but does not swap with it.
    Then the left neighbour $b$ of $a$ in $T$ is below $a$ in $R$ as well.
    Therefore, the tree $T'$ which can be received from $T$ by exchanging $a$ and $b$ must have caterpillar distance $d_c(T',R) = d_c(T,R) + 1 = l + 1$ to $R$.
    For an illustration of the $k^{th}$ loop when performing $\csort$ between $T$ and $R$ or $T'$ and $R$, respectively, consider Figure~\ref{fig:cat_convex_seq_example}.

    \todo{LC: replace this figure}
    \begin{figure}[H]
    	\centering
    	\includegraphics[width=\textwidth]{cat_convex_seq_example}
    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ as described in proof of Theorem~\ref{thm:caterpillar_convex}.
        The trees are represented as sequences that are being ordered by $\csort$.
        Right before loop number $k$, $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
    	\label{fig:cat_convex_seq_example}
    \end{figure}
